Ethnicity sensitive author disambiguation using semi-supervised learning

### 2016-01-19 paper/author_disambiguation.tex
extra '+'
solving an extra '+' which was added by mistake
Rephrase things

\section{Introduction}
In academic digital libraries, author name disambiguation is the problem of
grouping together publications written by the same person.
Author name disambiguation is often a difficult problem because an author may use different
spellings or name variants across their career (synonymy) and/or distinct authors may
share the same name (polysemy).
Most notably, author disambiguation is often more troublesome for researchers
from non-Western cultures, where personal names may be traditionally less diverse (leading to
homonym issues) or for which transliteration to Latin characters may not be unique (leading to
synonym issues).
With the fast growth of the scientific literature, author disambiguation has become a pressing
issue since the accuracy of information managed at the level of individuals directly affects:
the relevance search of results (\eg, when querying for all publications written by a given author);
the reliability of bibliometrics and author rankings (\eg, citation counts or other impact
metrics, as studied in \cite{strotmann2012author}); and/or the relevance of scientific network analysis
\cite{newman2001structure}. Thus, even small improvements in the field significantly improve the
+usability of the digital libraries to some users.

@@

As commonly performed in author disambiguation research,
we evaluate the predicted clusters over testing data  ${\cal C}_\text{test}^\prime$,
using both B3 and pairwise precision, recall
and F-measure, as defined below:
\begin{align}
P_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) &= \frac{1}{|{\cal S}|} \sum_{s \in {\cal S}} \frac{|c(s) \cap \widehat{c}(s)|}{|\widehat{c}(s)|} \\
R_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) &= \frac{1}{|{\cal S}|} \sum_{s \in {\cal S}} \frac{|c(s) \cap \widehat{c}(s)|}{|c(s)|}\\
F_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) &= \frac{2 P_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) R_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S})}{P_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) + P_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S})}\\
P_\text{pairwise}({\cal C}, \widehat{\cal C}) &= \frac{|p({\cal C}) \cap p(\widehat{\cal C})|}{|p(\widehat{\cal C})|}\\
R_\text{pairwise}({\cal C}, \widehat{\cal C}) &= \frac{|p({\cal C}) \cap p(\widehat{\cal C})|}{|p({\cal C})|}\\
F_\text{pairwise}({\cal C}, \widehat{\cal C}) &= \frac{2 P_\text{pairwise}({\cal C}, \widehat{\cal C}) R_\text{pairwise}({\cal C}, \widehat{\cal C})}{P_\text{pairwise}({\cal C}, \widehat{\cal C}) + R_\text{pairwise}({\cal C}, \widehat{\cal C})}
\end{align}
and where $c(s)$ (resp. $\widehat{c}(s)$) is the cluster $c \in {\cal C}$ such that
$s \in c$ (resp. the cluster $\widehat{c} \in \widehat{\cal C}$ such that $s
\in \widehat{c}$), and where $p({\cal C}) = \cup_{c \in {\cal C}} \{ (s_1, s_2)
| s_1, s_2 \in c, s_1 \neq s_2 \}$ is the set of all pairs of signatures from
the same clusters in ${\cal C}$.
Precision evaluates whether signatures are grouped only with signatures from the same true clusters,
while recall measures the extent to which all signatures from the same true clusters are
effectively grouped together.
The F-measure is the harmonic mean between these two quantities.
In the analysis below, we rely primarily on the B3 F-measure for discussing results, as the pairwise variant
tends to favor large clusters (because the number of pairs is quadratic with the cluster size),
hence unfairly giving preference to authors with many publications.
By contrast, the B3 F-measure weights clusters linearly with respect to their size.
General conclusions drawn below remain however consistent for pairwise F.

@@

\begin{table*}
\caption{Average precision, recall and f-measure scores on test folds.}
\label{table:results}
\centering
\begin{tabular}{|l|c c c | c c c|}
  \hline
                       & \multicolumn{3}{|c|}{\textbf{B3}} & \multicolumn{3}{|c|}{\textbf{Pairwise}}\\
  \textbf{Description} & $P$ & $R$ & $F$ & $P$ & $R$ & $F$ \\
  \hline
  \hline
Baseline & 0.9024 & 0.9828 & 0.9409  & 0.8298 & 0.9776 & 0.8977 \\
\hline
\underline{Blocking = SFI} & 0.9901 & 0.9760 & 0.9830  & 0.9948 & 0.9738 & 0.9842\\
Blocking = Double metaphone & 0.9856 & 0.9827 & 0.9841  & 0.9927 & 0.9817 & 0.9871 \\
Blocking = NYSIIS & 0.9875 & 0.9826 & \textbf{0.9850}  & 0.9936 & 0.9814 & \textbf{0.9875} \\
Blocking = Soundex & 0.9886 & 0.9745 & 0.9815 & 0.9935 & 0.9725 & 0.9828 \\
\hline
No name normalization & 0.9887 & 0.9697 & 0.9791 & 0.9931 & 0.9658 & 0.9793 \\
\underline{Name normalization} & 0.9901 & 0.9760 & \textbf{0.9830} & 0.9948 & 0.9738 & \textbf{0.9842}  \\
\hline
\hline
\underline{Classifier = GBRT} & 0.9901 & 0.9760 & 0.9830   & 0.9948 & 0.9738 & 0.9842 \\
Classifier = Random Forests & 0.9909 & 0.9783 & \textbf{0.9846}  & 0.9957 & 0.9752 & \textbf{0.9854} \\
Classifier = Linear Regression & 0.9749 & 0.9584 & 0.9666 & 0.9717 & 0.9569 & 0.9643 \\
\hline
Training pairs = Non-blocked, uniform & 0.9793 & 0.9630 & 0.9711  & 0.9756 & 0.9629 & 0.9692 \\
Training pairs = Blocked, uniform & 0.9854 & 0.9720 & 0.9786  & 0.9850 & 0.9707 & 0.9778 \\
\underline{Training pairs = Blocked, balanced} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
\hline
\underline{Clustering = Average linkage} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
Clustering = Single linkage & 0.9741 & 0.9603 & 0.9671  & 0.9543 & 0.9626 & 0.9584 \\
Clustering = Complete linkage & 0.9862 & 0.9709 & 0.9785  & 0.9920 & 0.9688 & 0.9803 \\
\hline
No cut (baseline) & 0.9024 & 0.9828 & 0.9409  & 0.8298 & 0.9776 & 0.8977 \\
Global cut & 0.9892 & 0.9737 & 0.9814  & 0.9940 & 0.9727 & 0.9832 \\
\underline{Block cut} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
\hline
Combined best settings & 0.9897 & 0.9827 & \textbf{0.9862}  & 0.9954 & 0.9805 & \textbf{0.9879} \\
  \hline
\end{tabular}
\end{table*}

Our approach could be considered as an experimental design solution. Our goal is to find the best combination of design parameters. Such a solution could be reached by performing many experiments where in each experiment we fix all design parameters except one. Of course, what we found as a best choice in each experiment will be used for the proceeding one and so on. To initialize the process, we will use what have been reported by some related works as a best method.

\textit{state-of-the-art.} Unfortunately, most methods proposed in related works have released neither their software, nor their data, making in practice a fair comparison very difficult. However, we believe state-of-the-art algorithms proposed in the literature could still be shown by categorizing the design parameters into the three categories; blocking strategy, linkage function, and clustering and compare the choices per each category. This categorization represents the three main phases in the pipeline. We believe that results for different settings somehow reflects some of these state-of-the-art choices as we will show in this section. Results presented in Table ~\ref{table:results} are discussed with
respect to a state-of-the-art using the following parameter design setting per category. In Table ~\ref{table:results}, what we considered as state-of-the-art choice is underlined which are as follows:

@@

\textit{Baseline.} Using machine learning techniques should perform better than dealing with the problem with other simpler method. Accordingly, we have set our experiment baseline having neither linkage function nor clustering. Actually, This is the case of many digital library where they use a simple and very fast execution solution with acceptable performance. Accordingly, our baseline algorithm is simply to group the set of publications from author names that have common surname and first initial as denoted in section \ref{methods:blocking} by \textit{SFI} and label them with a distinct author. However, this baseline also considers \textit{Name Normalization} for the 5 cases that is also previously discussed in section \ref{methods:blocking}. The results of the baseline is reported in the top of  Table ~\ref{table:results}.

In sections \ref{choices:blocking}, \ref{choices:linkage}, and \ref{choices:clustering}, we will discuss each group of design parameters in respect to the state-of-the-art choice of that group. Keep in mind that the results of each group uses the best of the other design parameters.

\subsubsection{Blocking choices} 
\label{choices:blocking}
The good precision of the state-of-the-art ($0.9901$), but its
lower recall ($0.9760$) suggest that the blocking strategy might be the
limiting factor to further overall improvements.
As shown in Table~\ref{table:blocking}, the maximum recall (\ie, if within a block, all signatures were clustered optimally) for SFI is $0.9828$.
At the price of fewer and therefore slightly larger blocks (as reported in the right column of Table~\ref{table:blocking}), the
proposed phonetic-based blocking strategies show better maximum recall (all
around $0.9905$), thereby pushing further the upper bound on the maximum
performance of author disambiguation.
Let us remind however that the reported maximum recalls for the blocking strategies using phonetization are
also raised due to the better handling of multiple surnames, as described in Section \ref{methods:blocking}.

@@

Finally, let us also note that Table~\ref{table:blocking} corroborates the
estimation of \cite{torvik2009author}, stating that SFI blocking has a recall
around $98\%$ on real data.

@@

\textit{Name normalization.} As discussed previously, the seemingly insignificant step of
normalizing author names (stripping accents, removing affixes), as performed in the
state-of-the-art, is shown to be important. Results from Table~\ref{table:results} clearly suggest that not
normalizing significantly reduces performance (yielding an F-measure of $0.9830$ when normalizing,
but decreasing to $0.9791$ when raw author name strings are used instead).

>>>>

\section{Introduction}
In academic digital libraries, author name disambiguation is the problem of
grouping together publications written by the same person.
Author name disambiguation is often a difficult problem because an author may use different
spellings or name variants across their career (synonymy) and/or distinct authors may
share the same name (polysemy).
Most notably, author disambiguation is often more troublesome for researchers
from non-Western cultures, where personal names may be traditionally less diverse (leading to
homonym issues) or for which transliteration to Latin characters may not be unique (leading to
synonym issues).
With the fast growth of the scientific literature, author disambiguation has become a pressing
issue since the accuracy of information managed at the level of individuals directly affects:
the relevance search of results (\eg, when querying for all publications written by a given author);
the reliability of bibliometrics and author rankings (\eg, citation counts or other impact
metrics, as studied in \cite{strotmann2012author}); and/or the relevance of scientific network analysis
\cite{newman2001structure}. Thus, even small improvements in the field significantly improve the usability of the digital libraries to some users.

@@

As commonly performed in author disambiguation research,
we evaluate the predicted clusters over testing data  ${\cal C}_\text{test}^\prime$,
using both B3 and pairwise precision, recall
and F-measure, as defined below:
\begin{align}
P_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) &= \frac{1}{|{\cal S}|} \sum_{s \in {\cal S}} \frac{|c(s) \cap \widehat{c}(s)|}{|\widehat{c}(s)|} \\
R_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) &= \frac{1}{|{\cal S}|} \sum_{s \in {\cal S}} \frac{|c(s) \cap \widehat{c}(s)|}{|c(s)|}\\
F_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) &= \frac{2 P_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) R_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S})}{P_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) + P_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S})}
\end{align}
\begin{align}
P_\text{pairwise}({\cal C}, \widehat{\cal C}) &= \frac{|p({\cal C}) \cap p(\widehat{\cal C})|}{|p(\widehat{\cal C})|}\\
R_\text{pairwise}({\cal C}, \widehat{\cal C}) &= \frac{|p({\cal C}) \cap p(\widehat{\cal C})|}{|p({\cal C})|}\\
F_\text{pairwise}({\cal C}, \widehat{\cal C}) &= \frac{2 P_\text{pairwise}({\cal C}, \widehat{\cal C}) R_\text{pairwise}({\cal C}, \widehat{\cal C})}{P_\text{pairwise}({\cal C}, \widehat{\cal C}) + R_\text{pairwise}({\cal C}, \widehat{\cal C})}
\end{align}
and where $c(s)$ (resp. $\widehat{c}(s)$) is the cluster $c \in {\cal C}$ such that
$s \in c$ (resp. the cluster $\widehat{c} \in \widehat{\cal C}$ such that $s
\in \widehat{c}$), and where $p({\cal C}) = \cup_{c \in {\cal C}} \{ (s_1, s_2)
| s_1, s_2 \in c, s_1 \neq s_2 \}$ is the set of all pairs of signatures from
the same clusters in ${\cal C}$.
Precision evaluates whether signatures are grouped only with signatures from the same true clusters,
while recall measures the extent to which all signatures from the same true clusters are
effectively grouped together.
The F-measure is the harmonic mean between these two quantities.
In the analysis below, we rely primarily on the B3 F-measure for discussing results, as the pairwise variant
tends to favor large clusters (because the number of pairs is quadratic with the cluster size),
hence unfairly giving preference to authors with many publications.
By contrast, the B3 F-measure weights clusters linearly with respect to their size.
General conclusions drawn below remain however consistent for pairwise F.

@@

\begin{table*}
\caption{Average precision, recall and f-measure scores on test folds. Underlined components correspond together the state-of-the-art solution.}
\label{table:results}
\centering
\begin{tabular}{|l|c c c | c c c|}
  \hline
                       & \multicolumn{3}{|c|}{\textbf{B3}} & \multicolumn{3}{|c|}{\textbf{Pairwise}}\\
  \textbf{Description} & $P$ & $R$ & $F$ & $P$ & $R$ & $F$ \\
  \hline
  \hline
Baseline & 0.9024 & 0.9828 & 0.9409  & 0.8298 & 0.9776 & 0.8977 \\
\hline
\underline{Blocking = SFI} & 0.9901 & 0.9760 & 0.9830  & 0.9948 & 0.9738 & 0.9842\\
Blocking = Double metaphone & 0.9856 & 0.9827 & 0.9841  & 0.9927 & 0.9817 & 0.9871 \\
Blocking = NYSIIS & 0.9875 & 0.9826 & \textbf{0.9850}  & 0.9936 & 0.9814 & \textbf{0.9875} \\
Blocking = Soundex & 0.9886 & 0.9745 & 0.9815 & 0.9935 & 0.9725 & 0.9828 \\
\hline
No name normalization & 0.9887 & 0.9697 & 0.9791 & 0.9931 & 0.9658 & 0.9793 \\
\underline{Name normalization} & 0.9901 & 0.9760 & \textbf{0.9830} & 0.9948 & 0.9738 & \textbf{0.9842}  \\
\hline
\hline
\underline{Classifier = GBRT} & 0.9901 & 0.9760 & 0.9830   & 0.9948 & 0.9738 & 0.9842 \\
Classifier = Random Forests & 0.9909 & 0.9783 & \textbf{0.9846}  & 0.9957 & 0.9752 & \textbf{0.9854} \\
Classifier = Linear Regression & 0.9749 & 0.9584 & 0.9666 & 0.9717 & 0.9569 & 0.9643 \\
\hline
Training pairs = Non-blocked, uniform & 0.9793 & 0.9630 & 0.9711  & 0.9756 & 0.9629 & 0.9692 \\
Training pairs = Blocked, uniform & 0.9854 & 0.9720 & 0.9786  & 0.9850 & 0.9707 & 0.9778 \\
\underline{Training pairs = Blocked, balanced} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
\hline
\underline{Clustering = Average linkage} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
Clustering = Single linkage & 0.9741 & 0.9603 & 0.9671  & 0.9543 & 0.9626 & 0.9584 \\
Clustering = Complete linkage & 0.9862 & 0.9709 & 0.9785  & 0.9920 & 0.9688 & 0.9803 \\
\hline
No cut (baseline) & 0.9024 & 0.9828 & 0.9409  & 0.8298 & 0.9776 & 0.8977 \\
Global cut & 0.9892 & 0.9737 & 0.9814  & 0.9940 & 0.9727 & 0.9832 \\
\underline{Block cut} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
\hline
Combined best settings & 0.9897 & 0.9827 & \textbf{0.9862}  & 0.9954 & 0.9805 & \textbf{0.9879} \\
  \hline
\end{tabular}
\end{table*}

% Our approach could be considered as an experimental design solution. Our goal is
% to find the best combination of design parameters. Such a solution could be
% reached by performing many experiments where in each experiment we fix all
% design parameters except one. Of course, what we found as a best choice in each
% experiment will be used for the proceeding one and so on. To initialize the
% process, we will use what have been reported by some related works as a best
% method.

\textit{Baseline.} The simplest baseline against which we compare our results
consists in grouping all signatures sharing the same (normalized) surname(s) and
the same (normalized) first given name initial. As it does not require the usage
of any machine learning component, this baseline provides a simple and fast
solution yielding decent results, as reported at the top of
Table~\ref{table:results}

\textit{State-of-the-art.} Most methods proposed in related works have released
neither their software, nor their data, making a fair comparison very difficult.
Yet, we believe solutions reported in the literature can be closely
matched to our generic pipeline, provided the blocking strategy, the linkage
function and the clustering algorithm are properly aligned. In particular,
we consider hereon as the \textit{state-of-the-art} solution the following
combination of components:

@@

Sections \ref{choices:blocking}, \ref{choices:linkage}, and
\ref{choices:clustering} now study each component individually and discuss
results with respect to the underlined state-of-the-art solution.

\subsubsection{Blocking choices}
\label{choices:blocking}

The good precision of the state-of-the-art ($0.9901$), but its
lower recall ($0.9760$) suggest that the blocking strategy might be the
limiting factor to further overall improvements.
As shown in Table~\ref{table:blocking}, the maximum recall (\ie, if within a block, all signatures were clustered optimally) for SFI is $0.9828$.
At the price of fewer and therefore slightly larger blocks (as reported in the right column of Table~\ref{table:blocking}), the
proposed phonetic-based blocking strategies show better maximum recall (all
around $0.9905$), thereby pushing further the upper bound on the maximum
performance of author disambiguation.
Let us remind however that the reported maximum recalls for the blocking strategies using phonetization are
also raised due to the better handling of multiple surnames, as described in Section \ref{methods:blocking}.

@@

Let us also note that Table~\ref{table:blocking} corroborates the
estimation of \cite{torvik2009author}, stating that SFI blocking has a recall
around $98\%$ on real data.

@@

Finally, as discussed previously, the seemingly insignificant step of
normalizing author names (stripping accents, removing affixes), as performed in the
state-of-the-art, is shown to be important. Results from Table~\ref{table:results} clearly suggest that not
normalizing significantly reduces performance (yielding an F-measure of $0.9830$ when normalizing,
but decreasing to $0.9791$ when raw author name strings are used instead).
